---
layout: post
title: "Sentiment Analysis with R"
subtitle: "Using rvest and Syuzhet."
author: "Chris Norval"
date: "2015-12-24 14:00:00"
output: 
  md_document:
    variant: markdown_github
header-img: "img/creative_2.jpg"
---

Sentiment analysis is a technique for evaluating the overall positivity, negativity, or neutrality of a body of text. It does this by scoring individual words based on whether they are positive or negative, and how strong a word it is in either direction. Positive words receive positive integers as a score, whereas negative words are represented by negative integers. The overall score for a body of text is the average of those numbers.


```{r, echo=F, message=F, warning=F}
library(dplyr) 
library(syuzhet)

sentList <- data.frame(Example = c(
  "good",
  "bad",
  "mediocre",
  "amazing",
  "good and bad",
  "amazing and bad"
), stringsAsFactors = F)

sentList$Score <- as.numeric(get_sentiment(sentList$Example))

colnames(sentList) <- c("Example Text", "Sentiment Score")

kable(sentList)
```

Sentiment analysis isn't perfect and there are plenty of examples where it will get things wrong, such as cases of sarcasm, context, or slang. For this reason, I'd be cautious using it for anything that requires rigour. 

That being said, it's an interesting technique for exploratory analysis. I've been looking for a good R package for sentiment analysis for quite some time, but more often than not they're depreciated or don't behave as expected. The `Syuzhet` package, however, seems to be what I was looking for. 

### Data

I decided to use an online transcript of the Leaders Debate before the UK's 2015 general election. I thought it'd be interesting to compare the sentiment of the different debaters and see how they vary over the course of the event. 

Firstly, load some packages and set up a few variables for graphing later.

```{r, message=F, warning=F}
library(rvest)      # Web scraping 

library(syuzhet)    # Sentiment package

library(dplyr)      # Wrangling

library(data.table) # Wrangling

library(zoo)        # For dragging down non-NA values to NA columns

library(ggplot2)    # Graphing



myTheme <- theme(legend.position="hidden", 
                 panel.background = element_blank(), 
                 panel.grid.major = element_line(colour = "grey"),
                 strip.background = element_rect(fill = "white"),
                 strip.text = element_text(face = "bold"))


colourList <- c("#999999", "#d50000", "#3F8428", "#008066", "#FDBB30", "#B3009D")
```

Then I use the `rvest` package to scrape the transcript and `dplyr` and `data.table` to reshape the dataset into a usable format. Feel free to skip over this block if you're only interested in the sentiment functions.

```{r, fig.width=7}

uri <- "http://webcache.googleusercontent.com/search?q=cache:yCrJUj1Pz-EJ:news-watch.co.uk/wp-content/uploads/2015/04/Transcript-of-BBC1-Opposition-Parties-Election-Debate.pdf+&cd=2&hl=en&ct=clnk&gl=uk"


# Read the html and split up

transcript <- read_html(uri) %>% 
  html_text() %>%
  gsub(":", ":\n", .) %>%
  strsplit(split = "\n") 


# Remove some of the html header content

transcript <- transcript[[1]][-(1:52)]


# Remove empty lines and page numbers

transcript <- data.table(speaker = as.character(NA), 
                         text = transcript, 
                         stringsAsFactors = F) %>%
  subset(text != "" & !(text %like% "Page "))


# Copy the speakers' names to their own column so that they can be dragged down.

transcript[text %in% c("DAVID DIMBLEBY:", "DD:"), speaker := "David Dimbleby"]
transcript[text %in% c("ED MILLIBAND:", "EM:"), speaker := "Ed Milliband"]
transcript[text %in% c("LEANNE WOOD:", "LW:", "LQ:"), speaker := "Leanne Wood"]
transcript[text %in% c("NIGEL FARAGE:", "NF:"), speaker := "Nigel Farage"]
transcript[text %in% c("NICOLA STURGEON:", "NS:"), speaker := "Nicola Sturgeon"]
transcript[text %in% c("NATALIE BENNETT:", "NB:"), speaker := "Natalie Bennett"]
transcript[text %in% c("EMILY MAITLIS:", "CHARLOTTE DENNIS:", "FIONA SWORD:", 
                       "ANAIA SHANANE:", "LEON MATTHEW:", 
                       "MICHAEL KENWORTHY:"), speaker := "Other"]


# Flag every row that contains unspoken dialogue as to be deleted.

transcript$toDelete <- F
transcript[!is.na(speaker), toDelete := T]


# Order the dialogue

transcript$order <- as.numeric(NA)
transcript[toDelete == T, order := as.numeric(seq(1, .N))]


# Drag down both the speaker and the order to NA columns

transcript$speaker <- na.locf(transcript$speaker)
transcript$order <- na.locf(transcript$order)


# Finally, delete the rows we're not interested in, group by order 

# (including speaker to retain that column), and paste the text together.

transcript <- transcript %>% 
  subset(speaker != "Other" & toDelete == F) %>%
  group_by(order, speaker) %>% 
  summarise(text = paste(text, collapse=" ")) %>%
  ungroup()

```

I tend to prefer `dplyr` over `data.table`, but you can't really go wrong knowing both. The `data.table` package allows you to easily assign a new value to a subset of your data based on a condition, as was done with speaker names in the previous step. 

So now we have the transcript in a more usable format, all we have to do to is call the get_sentiment function on the text column, and store the returning numbers in a new column.

```{r}

transcript$sentiment <- get_sentiment(transcript$text) %>% as.numeric()

```

### Analysis

Let's look at the min, max, mean and median sentiment scores for each speaker.

```{r, eval=F}

transcript %>%
  group_by(speaker) %>%
  summarise(min = min(sentiment), 
            max = max(sentiment), 
            mean = round(mean(sentiment), 2), 
            median = median(sentiment), 
            times_spoken = n())

```

```{r, echo=F}

transcript %>%
  group_by(speaker) %>%
  summarise(min = min(sentiment), 
            max = max(sentiment), 
            mean = round(mean(sentiment), 2), 
            median = median(sentiment), 
            times_spoken = n()) %>%
  kable

```

To find the most positive text, you can arrange by descending sentiment score and use head to return the first row:

```{r, eval=F}

transcript %>%
  arrange(desc(sentiment)) %>%
  select(speaker, sentiment, text) %>% 
  head(1)

```

```{r, echo=F}

transcript %>%
  arrange(desc(sentiment)) %>%
  select(speaker, sentiment, text) %>% 
  head(1) %>%
  kable

```

Alternatively, swap the `head` for `tail` to find the most negative speech. 

```{r, eval=F}

transcript %>%
  arrange(desc(sentiment)) %>%
  select(speaker, sentiment, text) %>% 
  tail(1)

```

```{r, echo=F}

transcript %>%
  arrange(desc(sentiment)) %>%
  select(speaker, sentiment, text) %>% 
  tail(1) %>%
  kable

```

It's worth pointing out that both the sentences with the lowest sentiment score and the highest sentiment score received applause from the audience. Therefore, just because text/speech has a low sentiment score doesn't mean that it will not resonate with people. Text with negative sentiment scores simply use a larger proportion of (or stronger) negative words. This is another example as to why, while interesting, you should be cautious about the assumptions you make when using this technique.

Using the order column as a rough proxy for time, we can also look at how each speaker's sentiment changes over the course of the debate. 

```{r 2015-12-13_sentiment_graph, fig.width=10, fig.height=7}

transcript %>%
  ggplot(aes(order, sentiment, color = speaker)) +
    geom_line(size = 1) + 
    scale_color_manual(values = colourList) +
    scale_x_continuous(name = "\nTime", breaks = NULL) +
    scale_y_continuous(name = "Sentiment\n", breaks = seq(-15, 15, by=5)) +
    facet_wrap(~ speaker) + 
    myTheme
    
```

### More Emotions

The `Syuzhet` package also supports the use of different Sentiment Analysis lexicons. The NRC method, for example, evaluates a body of text and returns positive values for eight different emotions (anger, anticipation, disgust, fear, joy, sadness, surprise, and trust), and two sentiments (negative and positive). You can compute these scores with the `get_nrc_sentiment()` function.

```{r 2015-12-13_sentiment_graph_2, fig.width=10, fig.height=9}

transcript <- cbind(transcript, get_nrc_sentiment(transcript$text))

transcript %>%
  gather("emotion", "score", 5:14) %>%
  ggplot(aes(order, score, color = speaker)) +
    geom_line(size = 1) + 
    scale_color_manual(values = colourList) +
    scale_x_continuous(name = "\nTime", breaks = NULL) +
    scale_y_continuous(name = "Sentiment\n", breaks = seq(-15, 15, by=5)) +
    facet_grid(speaker ~ emotion) + 
    myTheme

```

As the lexicons being used are different, it may be the case that the scores from the first method (afinn) and the scores from the second method (NRC) deviate from each other.

### Summary

The `Syuzhet` package is an easy way to get up and running with Sentiment Analysis on R. The different lexicons highlight that there is no one universally recognised technique, and no one technique is perfect. It can, however, lead to some interesting exploratory analysis, especially when combined with web scraping. I think it would be interesting to explore sentiment analysis as one input into a larger predictive model, and hopefully I'll get a chance to explore this over the next few weeks.





